---
title: |
       | \pkg{anytime}: Easier Date and Time Conversion

# Use letters for affiliations
author:
  - name: Dirk Eddelbuettel
    affiliation: 1
address:
  - code: 1
    address: Department of Statistics, University of Illinois, Urbana-Champaign, IL, USA

# For footer text  TODO(fold into template, allow free form two-authors)
lead_author_surname: Eddelbuettel

# Place DOI URL or CRAN Package URL here
doi: "https://cran.r-project.org/package=anytime"

# Abstract
abstract: |
  The \pkg{anytime} package converts from a number of different input
  variable types (integer, numeric, character, factor) and a number of
  different input formats which are tried heuristally offering a powerful
  and versatile date and time converter that (frequently) requires no
  user input.

# Font size of the document, values of 9pt (default), 10pt, 11pt and 12pt
fontsize: 9pt

# Optional: Force one-column layout, default is two-column
two_column: true

# Optional: Enables lineno mode, but only if one_column mode is also true
#lineno: true

# Optional: Enable one-sided layout, default is two-sided
#one_sided: true

# Optional: Enable section numbering, default is unnumbered
#numbersections: true

# Optional: Specify the depth of section number, default is 5
#secnumdepth: 5

# Optional: Skip inserting final break between acknowledgements, default is false
skip_final_break: true

# Optional: Bibliography
bibliography: anytime

# Optional: Enable a 'Draft' watermark on the document
watermark: false

# Customize footer, eg by referencing the vignette
footer_contents: "anytime Vignette"

# Produce a pinp document
output:
  pinp::pinp:
    collapse: true
    includes:
      in_header: header.tex

# Required: Vignette metadata for inclusion in a package.
vignette: >
  %\VignetteIndexEntry{anytime-introduction}
  %\VignetteKeywords{anytime, date, datetime, conversion}
  %\VignettePackage{anytime}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r initialsetup, include=FALSE}
knitr::opts_chunk$set(cache=FALSE)
```


# Motivation

R excels at computing with dates, and times.  Using _typed_
representation for your data is highly recommended not only because of
the functionality offered but also because of the added safety
stemming from proper representation.

But there is a small nuisance cost in interactive work as well as in
programming. Users must have told `as.POSIXct()` about a million times
that the origin is (of course) the
[epoch](https://en.wikipedia.org/wiki/Unix_time). Do we really have to
say it a million more times?  Similarly, when parsing dates that are
_some form_ of YYYYMMDD format, do we really have to manually convert
from `integer` or `numeric` or `factor` or `ordered` to character?
Having one of several common separators and/or date / time month forms
(YYYY-MM-DD, YYYY/MM/DD, YYYYMMDD, YYYY-mon-DD and so on, with or
without times), do we really need a format string? Or could a smart
converter function do this?

`anytime()` aims to be that _general purpose_ converter returning a
proper `POSIXct` (or `Date`) object no matter the input (provided it
was somewhat parseable), relying on
[Boost date_time](http://www.boost.org/doc/libs/1_61_0/doc/html/date_time.html)
for the (efficient, performant) conversion. `anydate()` is an
additional wrapper returning a `Date` object instead.

# Examples

We first set up the R environment and display for the examples that follow.

```{r, setup}
library(anytime)                 # our library
options(width=50,                # column width
        digits.secs=6)           # fractional secs
Sys.setenv(TZ=anytime:::getTZ()) # TZ helper
```


## From Integer or Numeric or Factor or Ordered

For the numeric date formats in the range of the (numeric) _yyyymmdd_, we use the `anydate()` function.

```{r, int_num_date}
## integer
anydate(20160101L + 0:2)

## numeric
anydate(20160101 + 0:2)
```

Numeric input also works for datetimes, but the range has to
correspond to the range of `as.numeric()` value of `POSIXct`
variables:

```{r, int_num_dt}
## integer
anytime(1451628000L + 0:2)

## numeric
anytime(1451628000 + 0:2)
```

This is a change in version 0.3.0; the old behaviour (which was not
fully consistent in how it treated numeric input values, but
convenient for input in the ranges shown here) can be enabled via
either an argument to the function or a global options, see
`help(anytime)` for details:

```{r, int_num_alt}
## integer
anytime(20160101L + 0:2, oldHeuristic=TRUE)

## numeric
anytime(20160101 + 0:2, oldHeuristic=TRUE)
```


## Factor or Ordered

Factor variables and their order variant are also supported directly.

```{r, fac_ord}
## factor
anytime(as.factor(20160101 + 0:2))

## ordered
anytime(as.ordered(20160101 + 0:2))
```

## Character: Simple

Character input is supported in a variety of formats.  We first show simple formats.

```{r, char_dates}
## Dates: Character
anytime(as.character(20160101 + 0:2))

## Dates: alternate formats
anytime(c("20160101", "2016/01/02", "2016-01-03"))
```


## Character: ISO

ISO8661 date (and datetime) formats are supported too.

```{r, char_iso}
## Datetime: ISO with/without fractional seconds
anytime(c("2016-01-01 10:11:12",
          "2016-01-01 10:11:12.345678"))
```

## Character: Textual month formats

Date formats with month abbreviations are supported in a number of common orderings.

```{r, char_date_text}
## ISO style
anytime(c("2016-Sep-01 10:11:12",
          "Sep/01/2016 10:11:12",
          "Sep-01-2016 10:11:12"))

## Datetime: Mixed format
## (cf http://stackoverflow.com/questions/39259184)
anytime(c("Thu Sep 01 10:11:12 2016",
          "Thu Sep 01 10:11:12.345678 2016"))
```

## Character: Dealing with DST

This shows an important aspect. When not working in localtime (by
overriding to `UTC`) the _changing difference_ UTC is correctly
covered (which the underlying
[Boost Date_Time](http://www.boost.org/doc/libs/1_61_0/doc/html/date_time.html)
library does not do by itself).


```{r, date_iso}
## Datetime: pre/post DST
anytime(c("2016-01-31 12:13:14",
          "2016-08-31 12:13:14"))
## important: catches change
anytime(c("2016-01-31 12:13:14",
          "2016-08-31 12:13:14"), tz="UTC")
```


# Technical Details

The actual parsing and conversion is done by two [Boost](http://www.boost.org) libraries.
First, [Boost
lexical_cast](http://www.boost.org/doc/libs/1_61_0/diaroc/html/boost_lexical_cast.html) is
used to convert from _anything_ to a string representation. This text representation is
then parsed by [Boost
Date_Time](http://www.boost.org/doc/libs/1_61_0/doc/html/date_time.html) to create the
corresponding date, or datetime, type.  (There are also a number of special cases where
numeric values are directly converted; see below.) We use the [BH
package](http://dirk.eddelbuettel.com/code/bh.html) \citep{CRAN:BH} to access these
[Boost](http://www.boost.org) libraries, and rely on
[Rcpp](http://dirk.eddelbuettel.com/code/rcpp.html) \citep{JSS:Rcpp,Eddelbuettel:2013} for
a seamless C++ interface to and from [R](https://www.r-project.org).

The [Boost Date_Time](http://www.boost.org/doc/libs/1_61_0/doc/html/date_time.html)
library is addressing the needs for parsing date and datetimes out of text. It permits us
to loop over a suitably large number of _candidate formats_ with considerable ease.  The
formats are generally variants of the ISO 8601 date format, _i.e._, of the YYYY-MM-DD
ordering.  In general, we also allow for textual representation of months, _e.g._, 'Jan'
for January.

```{r, date_single_digit}
anytime(c("2016-1-31", "2016-2-1"))
anytime(c("2016-1-31", "2016-2-1"), useR=TRUE)
```

The list of current formats can be retrieved by the `getFormats()`
function. Users can also add to this list at run-time by calling
`addFormats()`. User-provided formats are tried before the formats
supplied by the package.

```{r}
getFormats()
```

One minor shortcoming is that single digits dates and months are
sometimes encountered: 2001-2-3 for the third of February of 2001.
[Boost
Date_Time](http://www.boost.org/doc/libs/1_61_0/doc/html/date_time.html)
will only work with 2001-02-03 which is somewhat more restrictive.


As an alternative, and to be as close to parsing by the R language and system, we also
support the parser from R itself.  As R does not provide access via its API, we use the
RApiDatetime \citep{CRAN:RApiDatetime} package. This is shown in the second invocation of
`anytime()` in the preceding example.


# Output Formats

A related topic is faithful and easy to read _representation_ of
datetime objects in output, _i.e.,_ formats for printing object.

In the spirit of _no configuration_ used on the parsing side,
formating support is provided via several functions. These all follow
different known standards and are accessible by the name of the
standard, or, in one case, the non-standard convention. In all cases,
a character representation is returned.

```{r output_ex}
pt <- anytime("2016-01-31 12:13:14.123456")
iso8601(pt)
rfc2822(pt)
rfc3339(pt)
yyyymmdd(pt)
```

# TODOs

- [DONE] detail Boost use and parser
- [DONE] mention limits of that parser (no single digit day or month)
- [DONE] mention RApiDatetime use via useR=TRUE as well
- mention precision issue?
- [DONE] show formulae used?


<!-- ideally we would not need to force this here -->
\pagebreak
