---
title: |
       | \pkg{anytime}: Easier Date and Time Conversion

# Use letters for affiliations
author:
  - name: Dirk Eddelbuettel
    affiliation: a
address:
  - code: a
    address: Department of Statistics, University of Illinois, Urbana-Champaign, IL, USA

# For footer text  TODO(fold into template, allow free form two-authors)
lead_author_surname: Eddelbuettel

# Place DOI URL or CRAN Package URL here
doi: "https://cran.r-project.org/package=anytime"

# Abstract
abstract: |
  The \pkg{anytime} package converts from a number of different input
  variable types (integer, numeric, character, factor) and a number of
  different input formats which are tried heuristally offering a powerful
  and versatile date and time converter that (frequently) requires no
  user input.

# Font size of the document, values of 9pt (default), 10pt, 11pt and 12pt
fontsize: 9pt

# Optional: Force one-column layout, default is two-column
two_column: true

# Optional: Enables lineno mode, but only if one_column mode is also true
#lineno: true

# Optional: Enable one-sided layout, default is two-sided
#one_sided: true

# Optional: Enable section numbering, default is unnumbered
#numbersections: true

# Optional: Specify the depth of section number, default is 5
#secnumdepth: 5

# Optional: Skip inserting final break between acknowledgements, default is false
skip_final_break: true

# Optional: Bibliography
bibliography: anytime

# Optional: Enable a 'Draft' watermark on the document
watermark: true

# Customize footer, eg by referencing the vignette
footer_contents: "anytime Vignette"

# Produce a pinp document
output:
  pinp::pinp:
    includes:
      in_header: header.tex

# Required: Vignette metadata for inclusion in a package.
vignette: >
  %\VignetteIndexEntry{anytime-introduction}
  %\VignetteKeywords{anytime, date, datetime, conversion}
  %\VignettePackage{anytime}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r setup, include=FALSE}
#knitr::opts_chunk$set(collapse=TRUE)
knitr::opts_chunk$set(cache=TRUE)
options(width=55)
library(anytime)
Sys.setenv(TZ=anytime:::getTZ())  ## helper func
```


### Motivation

R excels at computing with dates, and times.  Using _typed_
representation for your data is highly recommended not only because of
the functionality offered but also because of the added safety
stemming from proper representation.

But there is a small nuisance cost in interactive work as well as in
programming. Users must have told `as.POSIXct()` about a million times
that the origin is (of course) the
[epoch](https://en.wikipedia.org/wiki/Unix_time). Do we really have to
say it a million more times?  Similarly, when parsing dates that are
_some form_ of YYYYMMDD format, do we really have to manually convert
from `integer` or `numeric` or `factor` or `ordered` to character?
Having one of several common separators and/or date / time month forms
(YYYY-MM-DD, YYYY/MM/DD, YYYYMMDD, YYYY-mon-DD and so on, with or
without times), do we really need a format string? Or could a smart
converter function do this?

`anytime()` aims to be that _general purpose_ converter returning a
proper `POSIXct` (or `Date`) object no matter the input (provided it
was somewhat parseable), relying on
[Boost date_time](http://www.boost.org/doc/libs/1_61_0/doc/html/date_time.html)
for the (efficient, performant) conversion. `anydate()` is an
additional wrapper returning a `Date` object instead.

### Examples

#### From Integer or Numeric or Factor or Ordered

```r
library(anytime)
options(digits.secs=6)           # fractional secs
Sys.setenv(TZ=anytime:::getTZ()) # TZ helper

## integer
anytime(20160101L + 0:2)
[1] "2016-01-01 CST" "2016-01-02 CST"
[3] "2016-01-03 CST"

## numeric
anytime(20160101 + 0:2)
[1] "2016-01-01 CST" "2016-01-02 CST"
[3] "2016-01-03 CST"

## factor
anytime(as.factor(20160101 + 0:2))
[1] "2016-01-01 CST" "2016-01-02 CST"
[3] "2016-01-03 CST"

## ordered
anytime(as.ordered(20160101 + 0:2))
[1] "2016-01-01 CST" "2016-01-02 CST"
[3] "2016-01-03 CST"
```

#### Character: Simple

```r
## Dates: Character
anytime(as.character(20160101 + 0:2))
[1] "2016-01-01 CST" "2016-01-02 CST"
[3] "2016-01-03 CST"

## Dates: alternate formats
anytime(c("20160101", "2016/01/02", "2016-01-03"))
[1] "2016-01-01 CST" "2016-01-02 CST"
[3] "2016-01-03 CST"
```


#### Character: ISO

```r
## Datetime: ISO with/without fractional seconds
anytime(c("2016-01-01 10:11:12",
          "2016-01-01 10:11:12.345678"))
[1] "2016-01-01 10:11:12.000000 CST"
[2] "2016-01-01 10:11:12.345678 CST"

## Datetime: ISO alternate (?) with 'T' separator
anytime(c("20160101T101112",
          "20160101T101112.345678"))
[1] "2016-01-01 10:11:12.000000 CST"
[2] "2016-01-01 10:11:12.345678 CST"
```

#### Character: Textual month formats

```r
## ISO style
anytime(c("2016-Sep-01 10:11:12",
          "Sep/01/2016 10:11:12",
          "Sep-01-2016 10:11:12"))
[1] "2016-09-01 10:11:12 CDT"
[2] "2016-09-01 10:11:12 CDT"
[3] "2016-09-01 10:11:12 CDT"

## Datetime: Mixed format
## (cf http://stackoverflow.com/questions/39259184)
anytime(c("Thu Sep 01 10:11:12 2016",
          "Thu Sep 01 10:11:12.345678 2016"))
[1] "2016-09-01 10:11:12.000000 CDT"
[2] "2016-09-01 10:11:12.345678 CDT"
```

#### Character: Dealing with DST

This shows an important aspect. When not working in localtime (by
overriding to `UTC`) the _changing difference_ UTC is correctly
covered (which the underlying
[Boost Date_Time](http://www.boost.org/doc/libs/1_61_0/doc/html/date_time.html)
library does not do by itself).


```r
## Datetime: pre/post DST
anytime(c("2016-01-31 12:13:14",
          "2016-08-31 12:13:14"))
[1] "2016-01-31 12:13:14 CST"
[2] "2016-08-31 12:13:14 CDT"
## important: catches change
anytime(c("2016-01-31 12:13:14",
          "2016-08-31 12:13:14"), tz="UTC")
[1] "2016-01-31 18:13:14 UTC"
[2] "2016-08-31 17:13:14 UTC"
```


### Technical Details

The heavy lifting is done by a combination of
[Boost lexical_cast](http://www.boost.org/doc/libs/1_61_0/doc/html/boost_lexical_cast.html)
to go from _anything_ to string representation which is then parsed by
[Boost Date_Time](http://www.boost.org/doc/libs/1_61_0/doc/html/date_time.html).
We use the [BH package](http://dirk.eddelbuettel.com/code/bh.html)
\citep{CRAN:BH} to access [Boost](http://www.boost.org), and rely on
[Rcpp](http://dirk.eddelbuettel.com/code/rcpp.html)
\citep{JSS:Rcpp,Eddelbuettel:2013} for a seamless C++ interface to
and from [R](https://www.r-project.org).


## TODOs

- detail Boost use and parser
- mention limits of that parser (no single digit day or month)
- mention RApiDatetime use via useR=TRUE as well
- figure out why codeblocks do not render collapsed in pinp


## Test

```{r foo, collapse=TRUE, results="markup"}
head(iris)
```
