---
title: \pkg{anytime}&#58; Easier Date and Time Conversion

# Use letters for affiliations
author:
  - name: Dirk Eddelbuettel
    affiliation: 1
address:
  - code: 1
    address: Department of Statistics, University of Illinois, Urbana-Champaign, IL, USA

# For footer text  TODO(fold into template, allow free form two-authors)
lead_author_surname: Eddelbuettel

# Place DOI URL or CRAN Package URL here
doi: "https://cran.r-project.org/package=anytime"

# Abstract
abstract: |
  The \pkg{anytime} package provides functions which convert from both a
  number of different input variable types (integer, numeric, character,
  factor) and different input formats which are tried heuristically offering
  a powerful and versatile date and time converter that (generally) requires
  no user input and operates autonomously.

# Font size of the document, values of 9pt (default), 10pt, 11pt and 12pt
fontsize: 9pt

# Optional: Force one-column layout, default is two-column
two_column: true

# Optional: Enables lineno mode, but only if one_column mode is also true
#lineno: true

# Optional: Enable one-sided layout, default is two-sided
#one_sided: true

# Optional: Enable section numbering, default is unnumbered
#numbersections: true

# Optional: Specify the depth of section number, default is 5
#secnumdepth: 5

# Optional: Skip inserting final break between acknowledgements, default is false
skip_final_break: true

# Optional: Bibliography
bibliography: anytime

# Optional: Enable a 'Draft' watermark on the document
watermark: false

# Customize footer, eg by referencing the vignette
footer_contents: "anytime Vignette"

# Produce a pinp document
output:
  pinp::pinp:
    collapse: true
    keep_tex: false

header-includes: >
  \newcommand{\proglang}[1]{\textsf{#1}}
  \newcommand{\pkg}[1]{\textbf{#1}}

# Required: Vignette metadata for inclusion in a package.
vignette: >
  %\VignetteIndexEntry{Introduction to anytime}
  %\VignetteKeywords{anytime, date, datetime, conversion}
  %\VignettePackage{anytime}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\usepackage[utf8]{inputenc}
---

```{r initialsetup, include=FALSE}
knitr::opts_chunk$set(cache=FALSE)
```


# Motivation

R excels at computing with dates, and times.  Using a _typed_ representation
for your data is highly recommended not only because of the functionality
offered but also because of the added safety stemming from proper
representation.

But there is a small nuisance cost in interactive work as well as in
programming. Users must have told `as.POSIXct()` about a million times that
the origin is (of course) the
[epoch](https://en.wikipedia.org/wiki/Unix_time). Do we really have to say it
a million more times?  Similarly, when parsing dates that are _some variant_
of the common YYYYMMDD format, do we really have to manually convert from
`integer` or `numeric` or `factor` or `ordered` to character?  Having one of
several common separators and/or date formats (YYYY-MM-DD, YYYY/MM/DD,
YYYYMMDD, YYYY-mon-DD and so on, with or without times), do we really need a
format string? Or could a smart converter function do this for us?

`anytime()` aims to be that _general purpose_ converter returning a proper
`POSIXct` (or `Date`) object no matter the input (provided it was parseable),
relying on [Boost
Date_Time](http://www.boost.org/doc/libs/1_69_0/doc/html/date_time.html) for
the (efficient, performant) conversion. `anydate()` is an additional wrapper
returning a `Date` object instead.


# Examples

We first set up the R environment and display for the examples that follow.

```{r, setup}
Sys.setenv(TZ=anytime:::getTZ()) # TZ helper
library(anytime)                 # our library, caches TZ info
options(width=50,                # column width
        digits.secs=6)           # fractional secs
```


## From Integer or Numeric or Factor or Ordered

For the numeric date formats in the range of the (numeric) _yyyymmdd_ format,
we use the `anydate()` function.

```{r, int_num_date}
## integer
anydate(20160101L + 0:2)

## numeric
anydate(20160101 + 0:2)
```

Numeric input also works for datetimes, but the range has to
correspond to the range of `as.numeric()` value of `POSIXct`
variables:

```{r, int_num_dt}
## integer
anytime(1451628000L + 0:2)

## numeric
anytime(1451628000 + 0:2)
```

This is a change from version 0.3.0; the old behaviour (which was not
fully consistent in how it treated numeric input values, but
convenient for input in the ranges shown here) can be enabled via
either an argument to the function or a global options, see
`help(anytime)` for details:

```{r, int_num_alt}
## integer
anytime(20160101L + 0:2, oldHeuristic=TRUE)

## numeric
anytime(20160101 + 0:2, oldHeuristic=TRUE)
```

## Factor or Ordered

Factor variables and their order variant are also supported directly.

```{r, fac_ord}
## factor
anytime(as.factor(20160101 + 0:2))

## ordered
anytime(as.ordered(20160101 + 0:2))
```

## Character: Simple

Character input is supported in a variety of formats.  We first show simple formats.

```{r, char_dates}
## Dates: Character
anytime(as.character(20160101 + 0:2))

## Dates: alternate formats
anytime(c("20160101", "2016/01/02", "2016-01-03"))
```


## Character: ISO

ISO8661 date (and datetime) formats are supported too, with both 'T' and a
space as separator of date and time.

```{r, char_iso}
## Datetime: ISO with/without fractional seconds
anytime(c("2016-01-01 10:11:12",
          "2016-01-01T10:11:12.345678"))
```

## Character: Textual month formats

Date formats with month abbreviations are supported in a number of common orderings.

```{r, char_date_text}
## ISO style
anytime(c("2016-Sep-01 10:11:12",
          "Sep/01/2016 10:11:12",
          "Sep-01-2016 10:11:12"))

## Datetime: Mixed format
## (cf http://stackoverflow.com/questions/39259184)
anytime(c("Thu Sep 01 10:11:12 2016",
          "Thu Sep 01 10:11:12.345678 2016"))
```

## Character: Dealing with DST

This shows an important aspect. When not working in localtime (by
overriding to `UTC`) the _change in difference_ to UTC is correctly
covered (which the underlying
[Boost Date_Time](https://www.boost.org/doc/libs/1_70_0/doc/html/date_time.html)
library does not do by itself).


```{r, date_iso}
## Datetime: pre/post DST
anytime(c("2016-01-31 12:13:14",
          "2016-08-31 12:13:14"))
## important: catches change
anytime(c("2016-01-31 12:13:14",
          "2016-08-31 12:13:14"), tz="UTC")
```

# Technical Details

The actual parsing and conversion is done by two different
[Boost](http://www.boost.org) libraries.  First, [Boost
lexical_cast](https://www.boost.org/doc/libs/1_69_0/doc/html/boost_lexical_cast.html)
is used to convert from _anything_ to a string representation.  This textual
representation is then parsed by [Boost
Date_Time](https://www.boost.org/doc/libs/1_69_0/doc/html/date_time.html) to
create the corresponding date, or datetime, type.  (There are also a number
of special cases where numeric values are directly converted; see below.) We
use the [\pkg{BH}](http://dirk.eddelbuettel.com/code/bh.html) package
\citep{CRAN:BH} to access these [Boost](https://www.boost.org) libraries, and
rely on [\pkg{Rcpp}](http://dirk.eddelbuettel.com/code/rcpp.html)
\citep{JSS:Rcpp,Eddelbuettel:2013,CRAN:Rcpp} for a seamless C++ interface to and from
[R](https://www.r-project.org).

The [Boost Date_Time](https://www.boost.org/doc/libs/1_69_0/doc/html/date_time.html)
library is addressing the needs for parsing date and datetimes from text. It permits us
to loop over a suitably large number of _candidate formats_ with considerable ease.  The
formats are generally variants of the ISO 8601 date format, _i.e._, of the YYYY-MM-DD
ordering.  In general, we also allow for textual representation of months, _e.g._, 'Jan'
for January.

```{r, date_single_digit}
anytime(c("2016-1-31", "2016-2-1"))
anytime(c("2016-1-31", "2016-2-1"), useR=TRUE)
```

The list of current formats can be retrieved by the `getFormats()`
function. Users can also add to this list at run-time by calling
`addFormats()`, as well as removing formats. User-provided formats are tried
before the formats supplied by the package.

```{r}
getFormats()
```

One minor shortcoming is that single digits dates and months are
sometimes encountered: 2001-2-3 for the third of February of 2001.
[Boost Date_Time](https://www.boost.org/doc/libs/1_68_0/doc/html/date_time.html)
will only work with 2001-02-03 which is somewhat more restrictive.

As an alternative, and to be as close to parsing by the R language and
system, we also support the parser from R itself.  As R does not provide
access via its API, we use the \pkg{Rcpp} package \citep{JSS:Rcpp,Eddelbuettel:2013,
CRAN:Rcpp}.  This is shown in the second invocation of `anytime()` in the
preceding example.

# Output Formats

A related topic is faithful and easy to read _representation_ of
datetime objects in output, _i.e.,_ formats for printing object.

In the spirit of _no configuration_ used on the parsing side,
formating support is provided via several functions. These all follow
different known standards and are accessible by the name of the
standard, or, in one case, the non-standard convention. In all cases,
a character representation is returned.

```{r output_ex}
pt <- anytime("2016-01-31 12:13:14.123456")
iso8601(pt)
rfc2822(pt)
rfc3339(pt)
yyyymmdd(pt)
```

# Ambiguities

The \pkg{anytime} package is designed to operate heuristically on a number of _plausible_
and _sane_ formats.  This cannot possibly cover all conceivable cases.

## North America versus the world

In general, \pkg{anytime} tries to gently nudge users towards ISO 8601 order of _year_
followed by _month_ and _day_. But for example in the United States, another prevalent
form insists on month-day-year ordering. As many users are likely to encounter such input
format, \pkg{anytime} accomodates this use provided a separator is used: either a slash
(`/`) or a hyphen (`-`) is accepted.

## One versus two digits

As mentioned earlier, displaying only one digit for day and/or month is possible. An
example would be "2010-2-4" for the fourth of February in 2010. However, the Boost parser
does not recognise this and would require "2010-02-04".  However, setting the option
`useR=TRUE` switches to the R parser which does recognise this input.

# Asserts

The \pkg{anytime} package also contains two helper functions that can assist in defensive
programming by validating input arguments. The `assertTime()` and `assertDate()` functions
validate if the given input can be parsed, respectively, as `Datetime` or `Date`
objects. In case one of the inputs cannot be parsed, an error is triggered.  Otherwise the
parsed input is returned invisibly.

# Summary

We describe the \pkg{anytime} package which offers fast, convenient and reliable date and
datetime conversion for R users along with helper functions for formatting and assertions.
